Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-12 analysis2_manuscript.tex
added pi sanity checks

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the sequential Markovian coalescent
    \citep{li2011inference,Schiffels2020,terhorst2017robust}, \kelcomment{mabye also cite Gutenkunst 2009 Dadi paper here} 
    %\igcomment{I'm not sure it's all that relevant, since it doesn't strictly infer N(t)}
    %\drscomment{One could use dadi for that, but then we should also cite things like fastsimcoal and ABC, while explaining
    % that users can use those methods to infer piecewise-constant N(t) histories as well with a prespecified number of epochs
    % and priors around their params, but maybe that is too much of a digression here? I'm fine either way}
    % ADK: I think we are good as is
    through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by descent information \citep{santiago2020recent}.
    Since all of these methods assume neutrality, they are typically applied to genomic
    segments that are not expected to be directly affected by selection (e.g., by masking exons
    and conserved elements).
    However, natural selection acting on linked sites---even those that are masked out of the
    analysis---can bias estimates of $N_e(t)$
    away from census population sizes by increasing the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}. 
    Using \stdpopsim, we can readily examine how different methods for inferring $N_e(t)$
    are influenced by selection on linked sites,
    which in the case of negative selection is commonly referred to as \emph{background selection}
    \citep{charlesworth1993effect,hudson1995deleterious}.

@@

    To do this we simulated human genomes with and without natural selection.
    In both scenarios, we ran three replicate simulations
    under the out of Africa (OOA) demographic model of
    \citep{ragsdale2019models} using a genetic map from the HapMap Project \citep{international2007second} (see \textbf{Methods}).
    Simulations with selection were implemented by modeling purifying selection on exons
    % IG: moved details to methods
    %, defined from the HAVANA group release 104 for the human genome,
    using a DFE inferred by \cite{kim2017inference}.
    Following the simulation, inference of $N_e(t)$ was conducted from each of the six simulated datasets using four methods: 1) \msmc \citep{Schiffels2020}, 
    2) \stairway \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    % IG: I commented this out because it appeared distracting. We can add a note in the end of this analysis
    % However, as a thorough benchmarking is not the main purpose of this manuscript, we saved computational resources by not conducting more simulation replicates.

@@

    To expand these observations to additional species and to highlight the ease of comparison enabled by \stdpopsim,
    we performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    We simulated 100 genomes under a two-epoch demographic model inferred for vaquita in \textcite{robinson2022critically},
    and a constant recombination rate across the genome (see \textbf{Methods}).
    In simulations with selection, we applied the DFE inferred by \textcite{robinson2022critically} to exons
    annotated from the vaquita genome assembly.
    This DFE model implements a relationship between selection and dominance 
    in which very deleterious mutations (with selection coefficient $s<-0.1$) are fully recessive
    (dominance coefficient $h=0$) and weakly deleterious mutations ($s\geq -0.001$) are nearly
    additive ($h=0.4$; see \textbf{Methods}).
    Here, we see that the site-frequency-based method (\stairway) produces more accurate
    inference than the two methods based on the sequential Markovian coalescent (\smcpp and \msmc; see Figure \ref{fig:vaquita-demography}).
    Reagrdless of these differences, we see that background selection again does not
    appear to considerably influence the inference of any of the four methods for this species,
    which has a very different evolutionary history than humans.

@@

    Overall, inferred DFEs varied across populations and inference methods (Figure \ref{fig:homsap-dfe.ooa}).
    Note that while the estimation accuracy for the DFE parameters themselves (Figure \ref{fig:homsap-dfe.ooa}A-B) is useful,
    examining the difference between the actual and inferred distributions can be more helpful for showing how well a method
    has recovered the true DFE (Figure \ref{fig:homsap-dfe.ooa}C).
    The distributions of $s$ inferred by \polydfe and \dadicli had higher mean values (in absolute value)
    but smaller shape parameters, while \grapes inferred distributions with lower mean values and shape parameters.
    As a result, \polydfe and \dadicli over-estimate the fraction of mutations under strong selection ($s<-0.1$),
    while \grapes considerably over-estimates the fraction of mutations under weak selection ($s\geq-0.0001$).
    % IG: commented this out to simplify the basic observation
    % and under-estimates the fraction of mutations with $s<-0.01$.

    Population history appears to influence different methods in slightly different ways.
    For instance, \dadicli and \grapes produce more accurate estimates from the YRI genomes 
    (relative to the other two populations), whereas \polydfe
    produces its most accurate estimate from the CHB genomes.
    To factor out the influence of demographic changes, we also analyzed genomes generated using similar
    simulations but with no population size changes (Figure \ref{fig:homsap-dfe.constant}).
    In this simpler setting, \polydfe overlapped the true mean selection coefficient and shape parameter, whereas 
    the other methods do not. 
    \grapes exhibits similar biases without population size changes as in the simulations with more complex demography,
    but with milder under-estimation of the shape parameter, but \dadicli shows an inverse pattern
    with overestimated shape parameter and slightly underestimated mean.

@@

    Next, we benchmarked these three DFE-inference methods on simulations of the vaquita porpoise genome.
    Recall that these simulations apply a gamma-distributed selection coefficient to mutations in exons
    with a relationship between the selection and dominance coefficients (see \textbf{Methods}).
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B). 
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}).
    We suspect that this is due to the fact that all DFE inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection. 
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the 
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous, 
    where their selective effect is $h s$ rather than $s$.

@@

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_all_sites_DFE_plot}
    \caption{
    \label{fig:homsap-dfe.constant.all_sites}
    Performance of methods to infer the distribution of fitness effects (DFE) from genetic data
    under constant population size  where all sites along the simulated chromosome can experience selected and neutral mutation
    as opposed to exonic regions only.\\
    \igcomment{This fig is not cited anywhere. Do we really need it?}
    }
\end{figure}

@@

\caption{\bf{Performance of DFE methods across the simulation scenarios}. 
Bold rows show the lowest mean absolute error (MAE) for the two DFE parameters
for each species.\\
\igcomment{This table is not cited anywhere. The analysis of ``all sites" is also not mentioned anywhere. Do we want to add it?}
}
\begin{tabular}{lllrrrrrr}
\toprule
species ID & demography & annotation & \makecell{MAE \\ $E|s|$ \\ dadi-cli} & \makecell{MAE \\ $E|s|$ \\ grapes} & \makecell{MAE \\ $E|s|$ \\ polyDFE} & \makecell{MAE \\ shape \\ dadi-cli} & \makecell{MAE \\ shape \\ grapes} & \makecell{MAE \\ shape \\ polyDFE} \\
\midrule
HomSap & Constant & all sites & 0.0027 & 0.0111 & \bf{0.0018} & \bf{0.00071} & 0.029 & 0.014 \\
HomSap & Constant & exons & \bf{0.0012} & 0.0101 & 0.0023 & 0.030 & 0.0086 & \bf{0.0068} \\
HomSap & OOA 5R19 & all sites & 0.012 & 0.012 & \bf{0.0074} & \bf{0.027} & 0.055 & 0.035 \\
HomSap & OOA 5R19 & exons & 0.014 & 0.0111 & \bf{0.0072} & 0.031 & 0.051 & \bf{0.024} \\
PhoSin & Constant & all sites & \bf{0.024} & 0.025 & 0.024 & 0.21 & 0.24 & \bf{0.19} \\
PhoSin & Constant & exons & 0.024 & 0.025 & 0.024 & \bf{0.23} & 0.25 & 0.23 \\
PhoSin & Vaquita2Epoch 1R22 & all sites & \bf{0.024} & 0.025 & 0.024 & 0.20 & 0.23 & \bf{0.18} \\
PhoSin & Vaquita2Epoch 1R22 & exons & 0.024 & 0.025 & \bf{0.024} & \bf{0.18} & 0.23 & 0.21 \\
\bottomrule
\end{tabular}
\label{tab:dfe_table}
\end{table}

>>>>

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the sequential Markovian coalescent
    \citep{li2011inference,Schiffels2020,terhorst2017robust}, 
    %\igcomment{I'm not sure it's all that relevant, since it doesn't strictly infer N(t)}
    %\drscomment{One could use dadi for that, but then we should also cite things like fastsimcoal and ABC, while explaining
    % that users can use those methods to infer piecewise-constant N(t) histories as well with a prespecified number of epochs
    % and priors around their params, but maybe that is too much of a digression here? I'm fine either way}
    % ADK: I think we are good as is
    through use of the site frequency spectrum (SFS) \citep{liu2020stairway},
    as well as through identity by descent information \citep{santiago2020recent}.
    Since all of these methods assume neutrality, they are typically applied to genomic
    segments that are not expected to be directly affected by selection (e.g., by masking exons
    and conserved elements).
    However, natural selection acting on linked sites---even those that are masked out of the
    analysis---can bias estimates of $N_e(t)$
    away from census population sizes by increasing the 
    rate of coalescence in the genome \citep[e.g.][]{schrider2016effects}. 
    Using \stdpopsim, we can readily examine how different methods for inferring $N_e(t)$
    are influenced by selection on linked sites,
    which in the case of negative selection is commonly referred to as \emph{background selection}
    \citep{charlesworth1993effect,hudson1995deleterious}.

@@

    To do this we simulated human genomes with and without natural selection.
    In both scenarios, we ran three replicate simulations
    under the out of Africa (OOA) demographic model of
    \citep{ragsdale2019models} using a genetic map from the HapMap Project \citep{international2007second} (see \textbf{Methods}).
    Simulations with selection were implemented by modeling purifying selection on exons
    % IG: moved details to methods
    %, defined from the HAVANA group release 104 for the human genome,
    using a DFE inferred by \cite{kim2017inference}.
    A basic summary of nucleotide diversity from these simulations is shown in Figure \ref{fig:homsap-summary-stats},
    which shows that the presence of selection decreases polymorphism in simulations with selection only slightly
    at the genome-wide level, but more substantially within exonic regions under both a constant population size
    and the out-of-Africa demographic model.
    Following the simulation, inference of $N_e(t)$ was conducted from each of the six simulated datasets using four methods: 1) \msmc \citep{Schiffels2020}, 
    2) \stairway \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    % IG: I commented this out because it appeared distracting. We can add a note in the end of this analysis
    % However, as a thorough benchmarking is not the main purpose of this manuscript, we saved computational resources by not conducting more simulation replicates.

@@

    To expand these observations to additional species and to highlight the ease of comparison enabled by \stdpopsim,
    we performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    We simulated 100 genomes under a two-epoch demographic model inferred for vaquita in \textcite{robinson2022critically},
    and a constant recombination rate across the genome (see \textbf{Methods}).
    In simulations with selection, we applied the DFE inferred by \textcite{robinson2022critically} to exons
    annotated from the vaquita genome assembly.
    This DFE model implements a relationship between selection and dominance 
    in which very deleterious mutations (with selection coefficient $s<-0.1$) are fully recessive
    (dominance coefficient $h=0$) and weakly deleterious mutations ($s\geq -0.001$) are nearly
    additive ($h=0.4$; see \textbf{Methods}).
    As above, a basic summary of nucleotide diversity from these simulations is shown in Figure \ref{fig:vaquita-summary-stats},
    which again shows that the presence of selection decreases polymorphism in simulations with selection, but that the effect
    is only readily apparent within exonic regions. Further in this case,
    we see that the two-epoch demographic model used in the simulations actually leads to a slight increase
    in diversity versus the constant population size model.

    In comparing demography inference methods, we see that the site-frequency-based method (\stairway) produces more accurate
    inference than the two methods based on the sequential Markovian coalescent (\smcpp and \msmc; see Figure \ref{fig:vaquita-demography}).
    Reagrdless of these differences, we see that background selection again does not
    appear to considerably influence the inference of any of the four methods for this species,
    which has a very different evolutionary history than humans.

@@

    Overall, inferred DFEs varied across populations and inference methods (Figure \ref{fig:homsap-dfe.ooa}, Table \ref{tab:dfe_table}).
    Note that while the estimation accuracy for the DFE parameters themselves (Figure \ref{fig:homsap-dfe.ooa}A-B) is useful,
    examining the difference between the actual and inferred distributions can be more helpful for showing how well a method
    has recovered the true DFE (Figure \ref{fig:homsap-dfe.ooa}C).
    The distributions of $s$ inferred by \polydfe and \dadicli had higher mean values (in absolute value)
    but smaller shape parameters, while \grapes inferred distributions with lower mean values and shape parameters.
    As a result, \polydfe and \dadicli over-estimate the fraction of mutations under strong selection ($s<-0.1$),
    while \grapes considerably over-estimates the fraction of mutations under weak selection ($s\geq-0.0001$).
    % IG: commented this out to simplify the basic observation
    % and under-estimates the fraction of mutations with $s<-0.01$.

    Population history appears to influence different methods in slightly different ways.
    For instance, \dadicli and \grapes produce more accurate estimates from the YRI genomes 
    (relative to the other two populations), whereas \polydfe
    produces its most accurate estimate from the CHB genomes.
    To factor out the influence of demographic changes, we also analyzed genomes generated using similar
    simulations but with no population size changes (Figure \ref{fig:homsap-dfe.constant}).
    In this simpler setting, \polydfe overlapped the true mean selection coefficient and shape parameter, whereas 
    the other methods do not. 
    \grapes exhibits similar biases without population size changes as in the simulations with more complex demography,
    but with milder under-estimation of the shape parameter, but \dadicli shows an inverse pattern
    with overestimated shape parameter and slightly underestimated mean. 
    As a sanity check, we also inferred the DFE from human genome simulations with a constant population size 
    (Figure \ref{fig:homsap-dfe.constant}, Table \ref{tab:dfe_table}), but largely the results are similar.

@@

    Next, we benchmarked these three DFE-inference methods on simulations of the vaquita porpoise genome.
    Recall that these simulations apply a gamma-distributed selection coefficient to mutations in exons
    with a relationship between the selection and dominance coefficients (see \textbf{Methods}).
    In this setting, all methods perform uniformly worse than in the human genome simulations,
    with consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter (Figure \ref{fig:vaquita-dfe}A-B, Table \ref{tab:dfe_table}). 
    Similar patterns are observed when re-running the simulations with a simpler demographic model
    without population size changes (Figure \ref{fig:vaquita-dfe.constant}, Table \ref{tab:dfe_table}).
    We suspect that this is due to the fact that all DFE inference methods used here assume an additive model ($h = \frac{1}{2}$),
    while mutations in the simulated genomes were more recessive ($h < \frac{1}{2}$), and their dominance decreased with the strength of selection. 
    To examine the influence of this model misspecification, we compared the inferred distributions of $s$ to the 
    simulated distribution of $2hs$, which would have been equal to $s$ under an additive model (Figure \ref{fig:vaquita-dfe}C).
    Reassuringly, all three methods infer distributions that appear to fit the main mode of the simulated distribution of $2hs$.
    This is likely because deleterious alleles are typically at low frequency and thus heterozygous, 
    where their selective effect is $h s$ rather than $s$.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/summary_stats.png}
    \caption{
    \label{fig:homsap-summary-stats}
    Summaries of diversity for human simulations. The left and right panels show
    nucleotide diversity ($\pi$) calculated over all three replicate datasets
    for the Constant size and Out-of-Africa demographic models respectively.
    Each panel shows four violins representing the distribution of $\pi$ values
    under a model without selection (blue), a model with background selection (orange),
    calculated from either the entire genome (labeled neutral or DFE)
    or from exonic regions only (labeled exons-only).
    }
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/summary_stats.png}
    \caption{
    \label{fig:vaquita-summary-stats}
    Summaries of diversity for vaquita simulations. The left and right panels show
    nucleotide diversity ($\pi$) calculated over all three replicate datasets
    for the Constant size and two-epoch demographic models respectively.
    Each panel shows four violins representing the distribution of $\pi$ values
    under a model without selection (blue), a model with background selection (orange),
    calculated from either the entire genome (labeled neutral or DFE)
    or from exonic regions only (labeled exons-only).
    }
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/HomSap/Constant/HomSap_Constant_Gamma_K17_all_sites_DFE_plot}
    \caption{
    \label{fig:homsap-dfe.constant.all_sites}
    Performance of methods to infer the distribution of fitness effects (DFE) from genetic data
    under constant population size  where all sites along the simulated chromosome can experience selected and neutral mutation
    as opposed to exonic regions only.\\
    }
\end{figure}

@@

\caption{\bf{Performance of DFE methods across the simulation scenarios}. 
Bold rows show the lowest mean absolute error (MAE) for the two DFE parameters
for each species.}

\begin{tabular}{lllrrrrrr}
\toprule
species ID & demography & annotation & \makecell{MAE \\ $E|s|$ \\ dadi-cli} & \makecell{MAE \\ $E|s|$ \\ grapes} & \makecell{MAE \\ $E|s|$ \\ polyDFE} & \makecell{MAE \\ shape \\ dadi-cli} & \makecell{MAE \\ shape \\ grapes} & \makecell{MAE \\ shape \\ polyDFE} \\
\midrule
%HomSap & Constant & all sites & 0.0027 & 0.0111 & \bf{0.0018} & \bf{0.00071} & 0.029 & 0.014 \\
HomSap & Constant & exons & \bf{0.0012} & 0.0101 & 0.0023 & 0.030 & 0.0086 & \bf{0.0068} \\
%HomSap & OOA 5R19 & all sites & 0.012 & 0.012 & \bf{0.0074} & \bf{0.027} & 0.055 & 0.035 \\
HomSap & OOA 5R19 & exons & 0.014 & 0.0111 & \bf{0.0072} & 0.031 & 0.051 & \bf{0.024} \\
%PhoSin & Constant & all sites & \bf{0.024} & 0.025 & 0.024 & 0.21 & 0.24 & \bf{0.19} \\
PhoSin & Constant & exons & 0.024 & 0.025 & 0.024 & \bf{0.23} & 0.25 & 0.23 \\
%PhoSin & Vaquita2Epoch 1R22 & all sites & \bf{0.024} & 0.025 & 0.024 & 0.20 & 0.23 & \bf{0.18} \\
PhoSin & Vaquita2Epoch 1R22 & exons & 0.024 & 0.025 & \bf{0.024} & \bf{0.18} & 0.23 & 0.21 \\
\bottomrule
\end{tabular}
\label{tab:dfe_table}
\end{table}
